---
title: "Assignment 3, Task 2: CA cetacean species richness"
author: "Catherine Takata"
date: "2/24/2021"
output: 
  html_document:
    theme: cerulean
    code_folding: show
    toc: true
---

## Summary 
In this task, we will create a raster of species richness of 35 cetacean species off the coast of California. We will present our results through a static map using ggplot with a coastline shapefile for refrence. 

Source: AquaMaps; Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S., Rees, T., & Froese, R. (2016)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster) 
library(tidyverse)
library(here)
library(sf)
library(fasterize)
library(rnaturalearth)
```

```{r, warning=F}
# Load in subsetted raster data 
# Use rasterstack() to read in multiple layers at once as a multi-layer 
ca_species <- list.files(path = here("ca_cetaceans"), pattern = "*.tif", full.names = TRUE)
species_stack <- raster::stack(ca_species)
```

## Set a probability threshold to determine presence of species 

```{r, warning = F}
# Set a probability threshold using a if else statement to reclassify the presence of species 
# If the species is greater than or equal to a 0.6 threshold, return 1 ~ "present" else return 0 ~ "absent" 
cetaceans_thresh <- function(x, thresh = .6) {
  y <- ifelse(x >= thresh, 1, NA)
  return(y)
}

# Use the calc() function to determine the species richness of the compounded raster stack 
# Determines how many species are "present" in each cell 
species_richness <- calc(species_stack, fun = cetaceans_thresh)
species_richness_sum <- calc(species_richness, fun = sum, na.rm = TRUE)
```

## Convert to a dataframe
Convert the raster data to a dataframe to plot with ggplot().  The `raster::rasterToPoints()` function converts a raster to a matrix of points - `x`, `y`, and `layer` (or the layer name if it's a named layer).  Then `as.data.frame()` can convert that matrix to a `data.frame` object suitable for `ggplot.`

```{r, warning=F}
species_richness_df <- raster::rasterToPoints(species_richness) %>%
  as.data.frame()
species_sum_df <- raster::rasterToPoints(species_richness_sum) %>%
  as.data.frame()
```

```{r, warning=F, message=F}
# Using rnaturalearth, obtain a pre-downloaded vector map for coastline data 
# Coastline data will add visual context to our raster data 
coastline <- ne_download(scale = 50, type = 'coastline', category = 'physical', returnclass = "sf")

# Use ggplot() to plot and populate with raster data + spatial data to produce a "map" 
ggplot() + 
  geom_raster(data = species_sum_df, aes(x = x, y = y, fill = layer)) +
  geom_sf(data = coastline) +
  coord_sf(xlim = c(-125, -115), ylim = c(32, 38)) + 
  scale_fill_gradient(low = 'white', high = 'red') +
  theme_void() +
  labs(title = "Species richness of cetacean species of CA",
       caption = "Figure 1")
```

**Figure 1**: Red shading indicates species richness of 35 cetacean species off the California coast. Probability of occurence was determined by choosing a probability threshold of 0.6 to indicate "presence" of each species. Increased saturation indicates higher density of species and more occurences. 



